### 2020-06-30 manuscript/manuscript.rmd
refined text on new analyses

For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling of to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices. Each of these was then analyzed using several models. We used ModelFinder to identify the

We tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQ-TREE (Extended Data Table 1). We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling (Extended Data Table 2). Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig. 3B; Supplementary Table 3).

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQ-TREE provides a new family of C60 models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared C60 results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices. Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large]. The Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and  IQ-TREE, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices. These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups when analyzed using Poisson-CAT. Thus, our results from reanalyses are largely similar to previous studies and indicate that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several representative data matrices.

### Drastic differences in phylogenetic signal between site- homogeneous and heterogeneous models

### Increase CAT categories lead to different topologies
The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix. Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix. Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQ-TREE results, potentially indicating that the C60 models might be already overfitting to most matrices (Supplementary Table 6). These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

@@

@@

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

>>>>

For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling of to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.

For each matrix, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] with BIC in IQ-TREE (except only LG+C60 models are used in 3 large matrices). In all cases, models with C60 fit these matrices better than the site-homogeneous models. We then inferred support under the best-fit model from IQ-TREE (Extended Data Table 2). We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling (Extended Data Table 2).

All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the ctenophore-sister hypothesis (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 3), the condiitions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two deifferent means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses that lead to support for Porifera-sister.

### Increase CAT categories lead to different topologies

There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (e.g., number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here. In published analyses, all these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes. Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.

A major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories across a phylogenetic data matrix, so the number of categories is variable. Models with C60 are set to 60 categories. Differences in analysis results could therefore be due to differences in the number of categories.

Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix. Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQ-TREE results, potentially indicating that the C60 models might be already overfitting to most matrices (Supplementary Table 6). These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

@@

### Differences in phylogenetic signal between site- homogeneous and heterogeneous models

@@

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.
